Shallow EDSLs and Object-Oriented Programming

### 2016-06-21 paper/sections/Interpretations.tex
More S3

Conversely, we can define multiple interpretations in a
modular and intuitive way with an OO language like Scala:
\lstinputlisting[linerange=2-15]{../src/wellsized/Circuit.scala}%APPLY:MULTIPLE_SCALA
Instead of modifying the original code, we define \lstinline{wellSized} modularly.
The encoding makes use of three OOP abstraction mechanisms:
\emph{inheritance}, \emph{subtyping} and \emph{type-refinement}.
Specifically, the new \lstinline{Circuit} is a subtype of
\lstinline{base.Circuit} and declares a new method \lstinline{wellSized}.
The hierarchy implements the new \lstinline{Circuit} by inheriting the corresponding class
from \lstinline{base} and
complementing the body of \lstinline{wellSized}.
Also, fields of \lstinline{Beside} are refined with the new \lstinline{Circuit}
type to avoid type mismatch in creating objects.

@@

Before giving an example of dependent interpretations, we first show how to add new
constructs. The new constructs are: \emph{stretch ns c} which inserts additional wires into the circuit \emph{c} by
summing up \emph{ns} and $above\ c_1\ c_2$ which combines two circuits of the same width vertically.
Their types are given below:
\lstinputlisting[linerange=39-40]{./code/shallowCircuit.hs}%APPLY:SYNTAX_TYPES

Adding new constructs is easy - just defining new classes that implement \lstinline{Circuit}:
\lstinputlisting[linerange=19-30]{../src/wellsized/Circuit.scala}%APPLY:VARIANT
Not arbitrary circuits can be combined using $above$ and $stretch$
as stated in their specifications.
We hence define the \texttt{wellSized} method for these two new constructs to
verify the constraints they imply:

Definitions of \texttt{wellSized} for the extended cases make it a dependent interpretation, as they
rely on another method defined on the \texttt{Circuit}, \texttt{width}, for calculating the width of a circuit.
Most importantly, our approach do not force dependent interpretation to be
defined together with what they depend on. In the case of \texttt{width} and
\texttt{wellSized}, they can be defined in separate families as long as the
signature of \texttt{width} is exposed to the definition of \texttt{wellSized}.

\paragraph{Solving Expression Problem} The code above shows how to
define new variant, combining with

\subsection{Context-sensitive Interpretations}
by declaring contexts as arguments of the method.
For space reasons, we omit the.

>>>>

Conversely, we can define introduce new interpretation in a
modular and intuitive way with an OO language like Scala:
\begin{figure}
\lstinputlisting[linerange=2-15]{../src/wellsized/Circuit.scala}%APPLY:MULTIPLE_SCALA
\caption{Adding new interpretations}
\label{code:operation}
\end{figure}
Instead of modifying the original code, we define \lstinline{wellSized} modularly.
The encoding makes use of three OOP abstraction mechanisms:
\emph{inheritance}, \emph{subtyping} and \emph{type-refinement}.
Specifically, the new \lstinline{Circuit} is a subtype of
\lstinline{base.Circuit} and declares a new method \lstinline{wellSized}.
The hierarchy implements the new \lstinline{Circuit} by inheriting the corresponding class
from \lstinline{base} and
complementing the body of \lstinline{wellSized}.
Also, fields of \lstinline{Beside} are refined with the new \lstinline{Circuit}
type to avoid type mismatch in creating objects.

@@

Before giving an example of dependent interpretations, we first show how to add new
constructs. The new constructs are: \emph{stretch ns c} which inserts additional wires into the circuit \emph{c} by
summing up \emph{ns} and $above\ c_1\ c_2$ which combines two circuits of the same width vertically.
\setlength{\grammarindent}{5em} % increase separation between LHS/RHS
The extended grammar is given below:
%\lstinputlisting[linerange=39-40]{./code/shallowCircuit.hs}%APPLY:SYNTAX_TYPES
\begin{grammar}
<circuit> ::= \ldots
\alt `above' <circuit> <circuit>
\alt `stretch' <positive-numbers> <circuit>
\end{grammar}

Figure~\ref{code:variant} shows how to add new constructs - just defining new classes that implement \lstinline{Circuit}.
\begin{figure}
\lstinputlisting[linerange=2-11]{../src/width/Variants.scala}%APPLY:VARIANT_SCALA
\caption{Adding new language constructs}
\label{code:variant}
\end{figure}
\paragraph{Solving the Expression Problem (EP)} As EDSLs evolve along the time, the
demand of new syntax and new semantics may arise. It would be good if we could
implement these features modularly. This, however, requires that the host
language should be capable of solving the EP. Figure~\ref{code:operation} and
Figure~\ref{code:variant} together illustrate how elegant the OO solution is.
In this sense, OO languages are good candidates for defining modular EDSLs.

Not arbitrary circuits can be combined using $above$ and $stretch$
as stated in their specifications.
We hence define the \lstinline{wellSized} method for these two new constructs to
verify the constraints they imply:
\lstinputlisting[linerange=2-11]{../src/wellsized/Variants.scala}%APPLY:DEPENDENT_SCALA
Definitions of \lstinline{wellSized} for the extended cases make it a dependent interpretation, as they
rely \lstinline{width} for calculating the width of a circuit.
What distinguishes the OO approach is that \lstinline{width} and \lstinline{wellSized}
can be defined separately. Essentially, with only the signature of
\lstinline{width} exposed to the \lstinline{Circuit} we can
define \lstinline{wellSized}.

\subsection{Context-sensitive Interpretations}
Unlike \lstinline{width} and \lstinline{wellSized} which can be defined with
only the given circuit, interpretations may need some mutable contexts for definition.
These contexts can be captured by arguments of the method. For example, a
function that collects all the connections between wires inside a circuit would have
the following signature:
\begin{lstlisting}
type Layout = List[List[Tuple2[Int,Int]]]
def tlayout(f: Int => Int): Layout
\end{lstlisting}
where the context \lstinline{f} may vary in recursive calls.
For space reasons, we omit the implementation details.

### 2016-06-21 paper/sections/Transformations.tex
Modifications according to the comment
More S3

\section{Transformations in Shallow Embedding}
Transformations are typically regarded as the privilege of deep embedding.
Since there is not a data structure persisting the AST, how to define transformations
becomes unclear, which becomes one of the main criticism of shallow embedding.
It is not completely true especially for the OO setting.
This section illustrates how to define various transformations in shallow embedding.
%In fact, transformations can still be defined in shallow embedding.

@@

Figure~\ref{code:desugar} shows how to implement desugaring in Scala.
\begin{figure}
\lstinputlisting[linerange=3-18]{../src/scala/desugar/Circuit.scala}%APPLY:DESUGAR_SCALA
\caption{OO implementation of desugaring}
\label{code:desugar}
\end{figure}
The newly introduced method \lstinline{desugar} in \lstinline{Circuit} returns a \lstinline{Circuit} reflecting that it is a transformation. The implementation of \lstinline{desugar} for each case is
straightforward: \emph{fan n} just returns itself; $beside\ c_1\ c_2$ recursively desugars $c_1$ and $c_2$ then wraps back
their results into an instance of \lstinline{Beside}; \emph{identity n} mimics the
formula.

%\weixin{define a show function in the and show the result after desugaring?}
Inspired by the Scala implementation, we can also define \lstinline{desugar} in Haskell:

\begin{figure}
\lstinputlisting[language=haskell,linerange=6-21]{./code/Desugar.hs}%APPLY:DESUGAR
\caption{Haskell implementation of desugaring}
\label{code:haskell}
\end{figure}
The definition relies on the laziness of Haskell.
However, different from the Scala version, introducing \lstinline{desugar} is done
through modifying the original code by inserting a new field in the record.
We will discuss modularity issues in detail in Section~\ref{}.

\subsection{Sophisticated Optimizations}
Sophisticated optimizations may need to inspect multiple internal
representations of values.
For example, we may want to merge consecutive vertical wires together:
$$
beside\ (identity\ m)\ (identity\ n) = identity\ (m + n)
$$
This rewrite rule applies only when both $c_1$ and $c_2$ of the $beside\ c_1\ c_2$
are $identity$, which has to inspect the internal representation of $c_1$ and $c_2$ simultaneously.
However, pattern matching can only be simulated only at the the top level, i.e. $beside$.
There is no way to tell what the class types of $c_1$ and $c_2$ are.
Fortunately, Scala has built-in support for pattern matching. Nevertheless, we
can not simply decorate classes with \lstinline{case} modifier as
it is not possible to have a case class that extends another case class.
Alternatively, we define the \lstinline{unapply} method, a.k.a \emph{extractor}, in the companion object
to tell the compiler how to destruct the object. Here comes the implementation
of the above optimization:
\lstinputlisting[linerange=2-20]{../src/scala/mergeIds/Circuit.scala}%APPLY:MERGEIDS_SCALA

>>>>

\section{Transformations in Shallow Embeddings}
Transformations are typically regarded as the privilege of deep embedding.
Since there is not a data structure persisting the AST, how to define transformations
becomes unclear, which becomes one of the main criticism of shallow embedding.
It is not completely true especially for the OO setting.
This section illustrates how to define various transformations in shallow embedding.
%In fact, transformations can still be defined in shallow embedding.

@@

Figure~\ref{code:desugar} shows how to implement desugaring in Scala.
\begin{figure}
\lstinputlisting[linerange=4-25]{../src/desugar/Circuit.scala}%APPLY:DESUGAR_SCALA
\caption{OO implementation of desugaring}
\label{code:desugar}
\end{figure}
The newly introduced method \lstinline{desugar} in \lstinline{Circuit} returns a \lstinline{Circuit} reflecting that it is a transformation. The implementation of \lstinline{desugar} for each case is
straightforward: \emph{fan n} just returns itself; $beside\ c_1\ c_2$ recursively desugars $c_1$ and $c_2$ then wraps back
their results into an instance of \lstinline{Beside}; \emph{identity n} mimics the
formula.

%\weixin{define a show function in the and show the result after desugaring?}
Inspired by the Scala implementation, we can port it back to Haskell:
\lstinputlisting[linerange=6-21]{./code/desugar.hs}%APPLY:DESUGAR
which relies on the laziness of Haskell.

\subsection{Sophisticated Optimizations}
Sophisticated optimizations may need to inspect multiple internal
representations of values.
Suppose we want to merge consecutive vertical wires together:
$$
beside\ (identity\ m)\ (identity\ n) = identity\ (m + n)
$$
This rewrite rule applies only when both $c_1$ and $c_2$ of the $beside\ c_1\ c_2$
are $identity$, which has to inspect the internal representation of $c_1$ and $c_2$ simultaneously.
However, pattern matching can only be simulated only at the the top level, i.e. $beside$.
There is no way to tell what the class types of $c_1$ and $c_2$ are.
Fortunately, Scala has built-in support for pattern matching. Nevertheless, we
can not simply decorate classes with \lstinline{case} modifier as
it is not possible to have a case class that extends another case class.
Alternatively, we define the \lstinline{unapply} method, a.k.a \emph{extractor}, in the companion object
to tell the compiler how to destruct the object. Here comes the implementation
of the above optimization:
\lstinputlisting[linerange=4-22]{../src/optimizations/Circuit.scala}%APPLY:MERGEIDS_SCALA
